---
title: "K-calibration"
author: "Farnaz Kamranzad"
date: "2023-03-08"
output: html_document
---


# Calling libraries
```{r, eval=TRUE, warning=FALSE, include=FALSE, echo=FALSE, message=FALSE}
start.time <- Sys.time()
knitr::opts_chunk$set(echo = TRUE)
library(tidyquant)
library(gridExtra)
library(grid)
library(lemon)
library(ggplot2)
library(inlabru)
library(INLA)
library(ETAS.inlabru)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(pracma)
library(Bolstad2)
```



# Calling my .R source file for assessing rate incompleteness
```{r}
# This source features modified functions for considering the rate incompleteness 
source('modified_functions.R')
```



# Making synthetic catalogue 
```{r}
# Defining 'true' ETAS parameters for generating a synthetic catalogue
mu <- 0.1
K <- 0.089
alpha <- 2.29
c <- 0.11
p <- 1.08
M0 <- 2.5
theta <- list(mu=mu, K=K, alpha=alpha, c=c, p=p)


# Simulating the synthetics
modelledDuration = 100 
Ht <- data.frame(ts=c(50), magnitudes=c(6.7)) 
samp.etas.list <- generate_temporal_ETAS_synthetic(theta=theta, beta.p=log(10), M0=M0, T1=0, T2=modelledDuration, Ht=Ht)
simulated.ETAS.catalogue <- bind_rows(samp.etas.list)
simulated.ETAS.catalogue <- simulated.ETAS.catalogue[order(simulated.ETAS.catalogue$ts),]
simulated.ETAS.catalogue$idx <- seq(1,length(simulated.ETAS.catalogue$ts))

simulated.ETAS.catalogue
```



# Defining an input list
```{r}
# A dataframe containing different starting values for the inversion algorithm
startingValues <- data.frame( mu=c(0.3),
                              K=c(0.1),
                              alpha=c(1.),
                              c=c(0.2),
                              p=c(1.01))


# Load a set of parameters that we will need to tweak for this application
fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
list.input <- create_input_list_temporal_noCatalogue(fpath)

# Tweak the variables laoded from the input file

M0_1 <- 1
M0_1.5 <- 1.5
M0_2 <- 2
M0_2.5 <- 2.5 #Original M0
M0_3 <- 3
M0_3.5 <- 3.5
M0_4 <- 4
M0_4.5 <- 4.5
M0_5 <- 5



list.input_1 <- list.input
list.input_1.5 <- list.input
list.input_2 <- list.input
list.input_2.5 <- list.input
list.input_3 <- list.input
list.input_3.5 <- list.input
list.input_4 <- list.input
list.input_4.5 <- list.input
list.input_5 <- list.input

```


```{r}

##########  M0_1  ############ 

list.input_1$M0 <- M0_1

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_1$mu.init <- startingValues$mu[i]
list.input_1$alpha.init <- startingValues$alpha[i]
list.input_1$K.init <- startingValues$K[i]
list.input_1$c.init <- startingValues$c[i]
list.input_1$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_1$th.init <- list(th.mu = inv_gamma_t(list.input_1$mu.init, list.input_1$a_mu, list.input_1$b_mu),
                           th.K = inv_loggaus_t(list.input_1$K.init, list.input_1$a_K, list.input_1$b_K),
                           th.alpha = inv_unif_t(list.input_1$alpha.init, list.input_1$a_alpha, list.input_1$b_alpha),
                           th.c = inv_unif_t(list.input_1$c.init, list.input_1$a_c, list.input_1$b_c),
                           th.p = inv_unif_t(list.input_1$p.init, list.input_1$a_p, list.input_1$b_p) )

startTime <- 0
list.input_1$time.int <- c(startTime, modelledDuration)
list.input_1$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_1$max_step)){
  list.input_1$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_1$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_1$th.init) # parameters initial values
  } else {
    list.input_1$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_1$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_1$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_1$th.init) # parameters initial values
    }


##########  M0_1.5  ############ 


list.input_1.5$M0 <- M0_1.5

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_1.5$mu.init <- startingValues$mu[i]
list.input_1.5$alpha.init <- startingValues$alpha[i]
list.input_1.5$K.init <- startingValues$K[i]
list.input_1.5$c.init <- startingValues$c[i]
list.input_1.5$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_1.5$th.init <- list(th.mu = inv_gamma_t(list.input_1.5$mu.init, list.input_1.5$a_mu, list.input_1.5$b_mu),
                           th.K = inv_loggaus_t(list.input_1.5$K.init, list.input_1.5$a_K, list.input_1.5$b_K),
                           th.alpha = inv_unif_t(list.input_1.5$alpha.init, list.input_1.5$a_alpha, list.input_1.5$b_alpha),
                           th.c = inv_unif_t(list.input_1.5$c.init, list.input_1.5$a_c, list.input_1.5$b_c),
                           th.p = inv_unif_t(list.input_1.5$p.init, list.input_1.5$a_p, list.input_1.5$b_p) )

startTime <- 0
list.input_1.5$time.int <- c(startTime, modelledDuration)
list.input_1.5$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_1.5$max_step)){
  list.input_1.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_1.5$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_1.5$th.init) # parameters initial values
  } else {
    list.input_1.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_1.5$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_1.5$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_1.5$th.init) # parameters initial values
    }


##########  M0_2  ############ 


list.input_2$M0 <- M0_2

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_2$mu.init <- startingValues$mu[i]
list.input_2$alpha.init <- startingValues$alpha[i]
list.input_2$K.init <- startingValues$K[i]
list.input_2$c.init <- startingValues$c[i]
list.input_2$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_2$th.init <- list(th.mu = inv_gamma_t(list.input_2$mu.init, list.input_2$a_mu, list.input_2$b_mu),
                           th.K = inv_loggaus_t(list.input_2$K.init, list.input_2$a_K, list.input_2$b_K),
                           th.alpha = inv_unif_t(list.input_2$alpha.init, list.input_2$a_alpha, list.input_2$b_alpha),
                           th.c = inv_unif_t(list.input_2$c.init, list.input_2$a_c, list.input_2$b_c),
                           th.p = inv_unif_t(list.input_2$p.init, list.input_2$a_p, list.input_2$b_p) )

startTime <- 0
list.input_2$time.int <- c(startTime, modelledDuration)
list.input_2$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_2$max_step)){
  list.input_2$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_2$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_2$th.init) # parameters initial values
  } else {
    list.input_2$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_2$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_2$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_2$th.init) # parameters initial values
    }


##########  M0_2.5  ############ 


list.input_2.5$M0 <- M0_2.5

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_2.5$mu.init <- startingValues$mu[i]
list.input_2.5$alpha.init <- startingValues$alpha[i]
list.input_2.5$K.init <- startingValues$K[i]
list.input_2.5$c.init <- startingValues$c[i]
list.input_2.5$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_2.5$th.init <- list(th.mu = inv_gamma_t(list.input_2.5$mu.init, list.input_2.5$a_mu, list.input_2.5$b_mu),
                           th.K = inv_loggaus_t(list.input_2.5$K.init, list.input_2.5$a_K, list.input_2.5$b_K),
                           th.alpha = inv_unif_t(list.input_2.5$alpha.init, list.input_2.5$a_alpha, list.input_2.5$b_alpha),
                           th.c = inv_unif_t(list.input_2.5$c.init, list.input_2.5$a_c, list.input_2.5$b_c),
                           th.p = inv_unif_t(list.input_2.5$p.init, list.input_2.5$a_p, list.input_2.5$b_p) )

startTime <- 0
list.input_2.5$time.int <- c(startTime, modelledDuration)
list.input_2.5$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_2.5$max_step)){
  list.input_2.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_2.5$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_2.5$th.init) # parameters initial values
  } else {
    list.input_2.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_2.5$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_2.5$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_2.5$th.init) # parameters initial values
    }


##########  M0_3  ############ 


list.input_3$M0 <- M0_3

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_3$mu.init <- startingValues$mu[i]
list.input_3$alpha.init <- startingValues$alpha[i]
list.input_3$K.init <- startingValues$K[i]
list.input_3$c.init <- startingValues$c[i]
list.input_3$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_3$th.init <- list(th.mu = inv_gamma_t(list.input_3$mu.init, list.input_3$a_mu, list.input_3$b_mu),
                           th.K = inv_loggaus_t(list.input_3$K.init, list.input_3$a_K, list.input_3$b_K),
                           th.alpha = inv_unif_t(list.input_3$alpha.init, list.input_3$a_alpha, list.input_3$b_alpha),
                           th.c = inv_unif_t(list.input_3$c.init, list.input_3$a_c, list.input_3$b_c),
                           th.p = inv_unif_t(list.input_3$p.init, list.input_3$a_p, list.input_3$b_p) )

startTime <- 0
list.input_3$time.int <- c(startTime, modelledDuration)
list.input_3$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_3$max_step)){
  list.input_3$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_3$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_3$th.init) # parameters initial values
  } else {
    list.input_3$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_3$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_3$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_3$th.init) # parameters initial values
    }


##########  M0_3.5  ############ 



list.input_3.5$M0 <- M0_3.5

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_3.5$mu.init <- startingValues$mu[i]
list.input_3.5$alpha.init <- startingValues$alpha[i]
list.input_3.5$K.init <- startingValues$K[i]
list.input_3.5$c.init <- startingValues$c[i]
list.input_3.5$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_3.5$th.init <- list(th.mu = inv_gamma_t(list.input_3.5$mu.init, list.input_3.5$a_mu, list.input_3.5$b_mu),
                           th.K = inv_loggaus_t(list.input_3.5$K.init, list.input_3.5$a_K, list.input_3.5$b_K),
                           th.alpha = inv_unif_t(list.input_3.5$alpha.init, list.input_3.5$a_alpha, list.input_3.5$b_alpha),
                           th.c = inv_unif_t(list.input_3.5$c.init, list.input_3.5$a_c, list.input_3.5$b_c),
                           th.p = inv_unif_t(list.input_3.5$p.init, list.input_3.5$a_p, list.input_3.5$b_p) )

startTime <- 0
list.input_3.5$time.int <- c(startTime, modelledDuration)
list.input_3.5$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_3.5$max_step)){
  list.input_3.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_3.5$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_3.5$th.init) # parameters initial values
  } else {
    list.input_3.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_3.5$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_3.5$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_3.5$th.init) # parameters initial values
    }


##########  M0_4  ############ 


list.input_4$M0 <- M0_4

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_4$mu.init <- startingValues$mu[i]
list.input_4$alpha.init <- startingValues$alpha[i]
list.input_4$K.init <- startingValues$K[i]
list.input_4$c.init <- startingValues$c[i]
list.input_4$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_4$th.init <- list(th.mu = inv_gamma_t(list.input_4$mu.init, list.input_4$a_mu, list.input_4$b_mu),
                           th.K = inv_loggaus_t(list.input_4$K.init, list.input_4$a_K, list.input_4$b_K),
                           th.alpha = inv_unif_t(list.input_4$alpha.init, list.input_4$a_alpha, list.input_4$b_alpha),
                           th.c = inv_unif_t(list.input_4$c.init, list.input_4$a_c, list.input_4$b_c),
                           th.p = inv_unif_t(list.input_4$p.init, list.input_4$a_p, list.input_4$b_p) )

startTime <- 0
list.input_4$time.int <- c(startTime, modelledDuration)
list.input_4$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_4$max_step)){
  list.input_4$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_4$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_4$th.init) # parameters initial values
  } else {
    list.input_4$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_4$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_4$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_4$th.init) # parameters initial values
    }


##########  M0_4.5  ############ 


list.input_4.5$M0 <- M0_4.5

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_4.5$mu.init <- startingValues$mu[i]
list.input_4.5$alpha.init <- startingValues$alpha[i]
list.input_4.5$K.init <- startingValues$K[i]
list.input_4.5$c.init <- startingValues$c[i]
list.input_4.5$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_4.5$th.init <- list(th.mu = inv_gamma_t(list.input_4.5$mu.init, list.input_4.5$a_mu, list.input_4.5$b_mu),
                           th.K = inv_loggaus_t(list.input_4.5$K.init, list.input_4.5$a_K, list.input_4.5$b_K),
                           th.alpha = inv_unif_t(list.input_4.5$alpha.init, list.input_4.5$a_alpha, list.input_4.5$b_alpha),
                           th.c = inv_unif_t(list.input_4.5$c.init, list.input_4.5$a_c, list.input_4.5$b_c),
                           th.p = inv_unif_t(list.input_4.5$p.init, list.input_4.5$a_p, list.input_4.5$b_p) )

startTime <- 0
list.input_4.5$time.int <- c(startTime, modelledDuration)
list.input_4.5$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_4.5$max_step)){
  list.input_4.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_4.5$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_4.5$th.init) # parameters initial values
  } else {
    list.input_4.5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_4.5$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_4.5$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_4.5$th.init) # parameters initial values
    }


##########  M0_5  ############ 


list.input_5$M0 <- M0_5

i=1
  
# Change the starting location, measured on the ETAS scale
list.input_5$mu.init <- startingValues$mu[i]
list.input_5$alpha.init <- startingValues$alpha[i]
list.input_5$K.init <- startingValues$K[i]
list.input_5$c.init <- startingValues$c[i]
list.input_5$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input_5$th.init <- list(th.mu = inv_gamma_t(list.input_5$mu.init, list.input_5$a_mu, list.input_5$b_mu),
                           th.K = inv_loggaus_t(list.input_5$K.init, list.input_5$a_K, list.input_5$b_K),
                           th.alpha = inv_unif_t(list.input_5$alpha.init, list.input_5$a_alpha, list.input_5$b_alpha),
                           th.c = inv_unif_t(list.input_5$c.init, list.input_5$a_c, list.input_5$b_c),
                           th.p = inv_unif_t(list.input_5$p.init, list.input_5$a_p, list.input_5$b_p) )

startTime <- 0
list.input_5$time.int <- c(startTime, modelledDuration)
list.input_5$T12 <- c(startTime, modelledDuration)
  

# Define options for inlabru 
if(is.null(list.input_5$max_step)){
  list.input_5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input_5$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input_5$th.init) # parameters initial values
  } else {
    list.input_5$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input_5$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input_5$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input_5$th.init) # parameters initial values
    }



```



# Attaching the simulated catalogue to the formatted input lists
```{r}

###### M0_1 ######
list.input_1$catalog <- simulated.ETAS.catalogue
list.input_1$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )

###### M0_1.5 ######
list.input_1.5$catalog <- simulated.ETAS.catalogue
list.input_1.5$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )

###### M0_2 ######
list.input_2$catalog <- simulated.ETAS.catalogue
list.input_2$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )


###### M0_2.5 ######
list.input_2.5$catalog <- simulated.ETAS.catalogue
list.input_2.5$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )


###### M0_3 ######
list.input_3$catalog <- simulated.ETAS.catalogue
list.input_3$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )


###### M0_3.5 ######
list.input_3.5$catalog <- simulated.ETAS.catalogue
list.input_3.5$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )



###### M0_4 ######
list.input_4$catalog <- simulated.ETAS.catalogue
list.input_4$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )


###### M0_4.5 ######
list.input_4.5$catalog <- simulated.ETAS.catalogue
list.input_4.5$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )


###### M0_5 ######
list.input_5$catalog <- simulated.ETAS.catalogue
list.input_5$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )

```


# Inversion: Fitting ETAS models according to the input lists
```{r}
ETAS.model.fit_M0_1   <- Temporal.ETAS.fit(list.input_1)
ETAS.model.fit_M0_1.5 <- Temporal.ETAS.fit(list.input_1.5)
ETAS.model.fit_M0_2   <- Temporal.ETAS.fit(list.input_2)
ETAS.model.fit_M0_2.5 <- Temporal.ETAS.fit(list.input_2.5)
ETAS.model.fit_M0_3   <- Temporal.ETAS.fit(list.input_3)
ETAS.model.fit_M0_3.5 <- Temporal.ETAS.fit(list.input_3.5)
ETAS.model.fit_M0_4   <- Temporal.ETAS.fit(list.input_4)
ETAS.model.fit_M0_4.5 <- Temporal.ETAS.fit(list.input_4.5)
ETAS.model.fit_M0_5   <- Temporal.ETAS.fit(list.input_5)
```




# Small bit of post processing: Attaching model outputs to the input lists
```{r}
i<-1
 
#####  M0_1  #####
list.output_M0_1 <-list()
list.output_M0_1[[i]] <- append(list.input_1, list(model.fit = ETAS.model.fit_M0_1))
list.output_M0_1[[i]]$runtime <- sum(list.output_M0_1[[i]]$model.fit$bru_timings$Time)
list.output_M0_1[[i]]$nEvents <- length(list.output_M0_1[[i]]$catalog[,1])

#####  M0_1.5  #####
list.output_M0_1.5 <-list()
list.output_M0_1.5[[i]] <- append(list.input_1.5, list(model.fit = ETAS.model.fit_M0_1.5))
list.output_M0_1.5[[i]]$runtime <- sum(list.output_M0_1.5[[i]]$model.fit$bru_timings$Time)
list.output_M0_1.5[[i]]$nEvents <- length(list.output_M0_1.5[[i]]$catalog[,1])


#####  M0_2  #####
list.output_M0_2 <-list()
list.output_M0_2[[i]] <- append(list.input_2, list(model.fit = ETAS.model.fit_M0_2))
list.output_M0_2[[i]]$runtime <- sum(list.output_M0_2[[i]]$model.fit$bru_timings$Time)
list.output_M0_2[[i]]$nEvents <- length(list.output_M0_2[[i]]$catalog[,1])


#####  M0_2.5  #####
list.output_M0_2.5 <-list()
list.output_M0_2.5 [[i]] <- append(list.input_2.5, list(model.fit = ETAS.model.fit_M0_2.5))
list.output_M0_2.5 [[i]]$runtime <- sum(list.output_M0_2.5 [[i]]$model.fit$bru_timings$Time)
list.output_M0_2.5 [[i]]$nEvents <- length(list.output_M0_2.5 [[i]]$catalog[,1])


#####  M0_3  #####
list.output_M0_3 <-list()
list.output_M0_3 [[i]] <- append(list.input_3, list(model.fit = ETAS.model.fit_M0_3))
list.output_M0_3 [[i]]$runtime <- sum(list.output_M0_3 [[i]]$model.fit$bru_timings$Time)
list.output_M0_3 [[i]]$nEvents <- length(list.output_M0_3 [[i]]$catalog[,1])


#####  M0_3.5  #####
list.output_M0_3.5 <-list()
list.output_M0_3.5 [[i]] <- append(list.input_3.5, list(model.fit = ETAS.model.fit_M0_3.5))
list.output_M0_3.5 [[i]]$runtime <- sum(list.output_M0_3.5 [[i]]$model.fit$bru_timings$Time)
list.output_M0_3.5 [[i]]$nEvents <- length(list.output_M0_3.5 [[i]]$catalog[,1])


#####  M0_4  #####
list.output_M0_4 <-list()
list.output_M0_4 [[i]] <- append(list.input_4, list(model.fit = ETAS.model.fit_M0_4))
list.output_M0_4 [[i]]$runtime <- sum(list.output_M0_4 [[i]]$model.fit$bru_timings$Time)
list.output_M0_4 [[i]]$nEvents <- length(list.output_M0_4 [[i]]$catalog[,1])


#####  M0_4.5  #####
list.output_M0_4.5 <-list()
list.output_M0_4.5 [[i]] <- append(list.input_4.5, list(model.fit = ETAS.model.fit_M0_4.5))
list.output_M0_4.5 [[i]]$runtime <- sum(list.output_M0_4.5 [[i]]$model.fit$bru_timings$Time)
list.output_M0_4.5 [[i]]$nEvents <- length(list.output_M0_4.5 [[i]]$catalog[,1])


#####  M0_5  #####
list.output_M0_5 <-list()
list.output_M0_5 [[i]] <- append(list.input_5, list(model.fit = ETAS.model.fit_M0_5))
list.output_M0_5 [[i]]$runtime <- sum(list.output_M0_5 [[i]]$model.fit$bru_timings$Time)
list.output_M0_5 [[i]]$nEvents <- length(list.output_M0_5 [[i]]$catalog[,1])

```



# Posterior analysis
```{r}
plots <- list()

trueParas <- data.frame(value=c(mu, K, alpha,c, p), param=c("mu", "K", "alpha","c", "p"))

post.list_M0_1 <- get_posterior_param(input.list= list.output_M0_1[[1]])
post.df_M0_1 <-  post.list_M0_1[[1]]
post.df_M0_1$id <- "M0_1"

post.list_M0_1.5 <- get_posterior_param(input.list= list.output_M0_1.5[[1]])
post.df_M0_1.5 <-  post.list_M0_1.5[[1]]
post.df_M0_1.5$id <- "M0_1.5"

post.list_M0_2 <- get_posterior_param(input.list= list.output_M0_2[[1]])
post.df_M0_2 <-  post.list_M0_2[[1]]
post.df_M0_2$id <- "M0_2"

post.list_M0_2.5 <- get_posterior_param(input.list= list.output_M0_2.5[[1]])
post.df_M0_2.5 <-  post.list_M0_2.5[[1]]
post.df_M0_2.5$id <- "M0_2.5"

post.list_M0_3 <- get_posterior_param(input.list= list.output_M0_3[[1]])
post.df_M0_3 <-  post.list_M0_3[[1]]
post.df_M0_3$id <- "M0_3"

post.list_M0_3.5 <- get_posterior_param(input.list= list.output_M0_3.5[[1]])
post.df_M0_3.5 <-  post.list_M0_3.5[[1]]
post.df_M0_3.5$id <- "M0_3.5"

post.list_M0_4 <- get_posterior_param(input.list= list.output_M0_4[[1]])
post.df_M0_4 <-  post.list_M0_4[[1]]
post.df_M0_4$id <- "M0_4"

post.list_M0_4.5 <- get_posterior_param(input.list= list.output_M0_4.5[[1]])
post.df_M0_4.5 <-  post.list_M0_4.5[[1]]
post.df_M0_4.5$id <- "M0_4.5"

post.list_M0_5 <- get_posterior_param(input.list= list.output_M0_5[[1]])
post.df_M0_5 <-  post.list_M0_5[[1]]
post.df_M0_5$id <- "M0_5"

```



```{r, fig.height=4, fig.width=15}

ggplot(NULL, aes(x, y)) + 
  geom_line(data=post.df_M0_1, aes(group=id, color='M0_1'), size=1)+
  geom_line(data=post.df_M0_1.5, aes(group=id, color='M0_1.5'), size=1)+
  geom_line(data=post.df_M0_2, aes(group=id, color='M0_2'), size=1)+
  geom_line(data=post.df_M0_2.5, aes(group=id, color='M0_2.5'), size=1)+
  geom_line(data=post.df_M0_3, aes(group=id, color='M0_3'), size=1)+
  geom_line(data=post.df_M0_3.5, aes(group=id, color='M0_3.5'), size=1)+
  geom_line(data=post.df_M0_4, aes(group=id, color='M0_4'), size=1)+
  geom_line(data=post.df_M0_4.5, aes(group=id, color='M0_4.5'), size=1)+
  geom_line(data=post.df_M0_5, aes(group=id, color='M0_5'), size=1)+
  facet_wrap(facets = vars(param), scales = 'free', labeller = label_parsed, nrow=1) +
  geom_vline(aes(xintercept = value ), data=trueParas, color='black', linetype=2, label="True value") + 
  labs(color='Catalogue scenario', linetype='Catalogue scenario' ) +
  ggtitle(paste("calibration of K" )) +
  xlab("ETAS Posteriors") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
  #theme(legend.position='hidden')

```



# Computation time
```{r}
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```







