---
title: "Testing the history match"
author: "Naylor"
date: "2022-10-27"
output: html_document
---

# Analysis with variable duration of relatively quiet background period prior to M6.7 mainshock

Here we explore the robustness of the posteriors to varying the duration of a quiet period prior to a M6.7 mainshock. The hypothesis is that a representative sample needs to contain seperate periods that are dominated by the triggering behaviour and the background behaviour if the inversion is to work well.

In order to keep a similar number of events between the simulations, we generate a 1000 day catalogue and seed a M6.7 event on day 500.1 such that the early part of the catalogue is relativly quiet and the latter part contains an active sequence. In this baseline case, the posteriors are returned accurately. We then progressively remove the early part of the quiet period and explore how the posteriors vary. If the inversion is inensitive to having a quiecent period - the posteriors will consistently return the true solutions.

## Load some packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ETAS.inlabru)

library(tidyquant)
library(gridExtra)
library(grid)
library(lemon)
library(ggplot2)

library(inlabru)
library(INLA)

library(dplyr)

```

## Define the parameters for the syntheirc catalogues and starting values for inversion

```{r Create synthetic catalogue and plot}

# Parameters we use to generate synthetics, which we will refer to as the 'true' parameters
mu <- 0.1
K <- 0.089
alpha <- 2.29
c <- 0.11
p <- 1.08

# Format the true ETAS parameters for code to generate the synthetics
theta_etas <- data.frame(mu=mu, K=K, alpha=alpha, c=c, p=p)

# A dataframe containing different starting values for the algorithm
## This format was used when varying the starting conditions, but we only have one case here
startingValues <- data.frame( mu=c( 0.3),
                              K=c( 0.1),
                              alpha=c( 1.),
                              c=c( 0.2),
                              p=c( 1.01))

nRealisations <- length(startingValues$mu)


# Temporal duration of the synthetic catalogue in days
modelledDuration = 1000  # [days]

# The minimum magnitude that will be modelled
M0 = 2.5

```

## Generate new catalogues and save them

Fistly, generate a 1000 day catalogue with an M6.7 event seeded on day 500.1.

```{r}

############
#### Generate the second catalogue with a M6.7 event on day 500.01

Ht <- data.frame(ts=c(500.01), magnitudes=c(6.7))   # Impose a M6.7 event on day 500.01
samp.etas.list <- generate.temporal.ETAS.synthetic(theta = theta_etas, beta.p = log(10), M0 = M0, 
                                  T1 = 0, T2 = modelledDuration, Ht=Ht)

M6p7.ETAS.cat <- bind_rows(samp.etas.list)
M6p7.ETAS.cat <- M6p7.ETAS.cat[order(M6p7.ETAS.cat$ts),]

```

Now subset the catalogue using the start time list to progressively remove the first $startTime$ days off the beginning of the baseline catalogue. I could have implemented this just by changing $T1$, but instead I have cropped the data off the catalogues and shifted the origin back to 0. It is important to ensure that the model duation corresponds when setting up the inversion.

The list of the new catalogues is stored in $catList$.

```{r}

startTimes <- c(0,250,400,500,501)
catList <- list()

for (i in 1:length(startTimes)){
  startTime <- startTimes[i]
  tmp <- data.frame( M6p7.ETAS.cat[M6p7.ETAS.cat$ts>=startTime, ] )
  tmp$ts <- tmp$ts - startTime
  catList[[i]] <- tmp
}

```

Save the list of catalogues.

```{r}

save(catList,file="catList.Rda")

```

# Plot a summary of the catalogues

## Load the catalogues

```{r}

load("catList.Rda")

```

## Plot properties of the catalogues

```{r fig.height=1, fig.width=4}
#plots <-list()

df_start <- data.frame(startTimes<-startTimes)
df_large <- data.frame(x=c(500), y=c(6.7))
catPlot<- ggplot() + 
  geom_point(data = catList[[1]], aes(x=ts, y=magnitudes), size=0.2, alpha=0.5) +
  xlim(0, 1000) +
  ylim(2,7) +
  xlab("Time [days]") +
  ylab("Magnitude") +
    theme_bw()+
  geom_vline(data=df_start, aes( xintercept=startTimes, color=factor(startTimes), linetype=factor(startTimes))) +
  geom_point(data=df_large, aes(x=x, y=y), size=1, color='black')+
  ggtitle("(a) Variable start dates for inversion")+ 
  labs(color='Start time [days]', linetype='Start time [days]' )


#marrangeGrob(plots, nrow=length(catList), ncol=1)

#ggsave("initialConditionCats_varyRunin.png", marrangeGrob(plots, nrow=2, ncol=1))
```

```{r fig.height=6, fig.width=4}
plots <-list()


for(i in 1:length(catList)){
  
plots[[i]] <- ggplot() + 
  geom_point(data = catList[[i]], aes(x=ts, y=magnitudes), size=0.1, alpha=0.5) +
  xlim(0, 1000) +
  ggtitle(paste(500-startTimes[i], "days of runin, nEvents =", length(catList[[i]]$ts) )) +
  ylim(2,7) +
  xlab("Time [days]") +
  ylab("Magnitude") +
  theme_bw() +
  geom_vline( xintercept=1000-startTimes[i])
}

marrangeGrob(plots, nrow=length(catList), ncol=1)

#ggsave("initialConditionCats_varyRunin.png", marrangeGrob(plots, nrow=2, ncol=1))
```

# Analyse the robustness of the posteriors to quiecent runin period

## Loop over the catalog list and invert for the posteriors for each catalog

-   Loop over catalogues
-   Make sure that the period of the inversion is correct for the variable size catalogues
-   Store the results in `list.output.quietRuninScenario`

```{r}
## This code creates a list for the version conditioned on the history

list.output.quietRuninScenario_V2 <-list()

  # Load a set of parameters that we will need to tweak for this application
#fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
#list.input <- create.input.list.temporal.noCatalogue(fpath)
fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
list.input <- create.input.list.temporal.noCatalogue(fpath)
  
  
  ####################
  # Tweak the variables laoded from the input file
list.input$M0 <- M0

i=1
  
  # Change the starting location, measured on the ETAS scale
list.input$mu.init <- startingValues$mu[i]
list.input$alpha.init <- startingValues$alpha[i]
list.input$K.init <- startingValues$K[i]
list.input$c.init <- startingValues$c[i]
list.input$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma.t(x, a_mu, b_mu), 
             K = \(x) loggaus.t(x, a_K, b_K), 
             alpha = \(x) unif.t(x, a_alpha, b_alpha), 
             c_ = \(x) unif.t(x, a_c, b_c), 
             p = \(x) unif.t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input$th.init <- list(th.mu = inv.gamma.t(list.input$mu.init, list.input$a_mu, list.input$b_mu),
                th.K = inv.loggaus.t(list.input$K.init, list.input$a_K, list.input$b_K),
                th.alpha = inv.unif.t(list.input$alpha.init, list.input$a_alpha, list.input$b_alpha),
                th.c = inv.unif.t(list.input$c.init, list.input$a_c, list.input$b_c),
                th.p = inv.unif.t(list.input$p.init, list.input$a_p, list.input$b_p) )


for (i in 1:length(startTimes)) {  
  startTime <- startTimes[i]
  list.input$time.int <- c(startTime, modelledDuration)
  list.input$T12 <- c(startTime, modelledDuration)
  
    # Define options for inlabru 
  if(is.null(list.input$max_step)){
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         num.threads = 5,
                         #bru_method = list(max_step = 0.5),
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  } else {
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         bru_method = list(max_step = list.input$max_step),
                         num.threads = 5,
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  }

  ## Add out catalogue to the input list
  list.input$catalog <- catList[[1]]
  
  ## Add the catalogue formatted for bru
  list.input$catalog.bru <- data.frame(
    ts = catList[[1]]$ts,
    magnitudes = catList[[1]]$magnitudes,
    idx.p = seq_len(nrow(catList[[1]]))
  )
  
  ## Input list is now formatted
  ####################  
  
  ## Run the model according to the input list
  ETAS.model.fit <- Temporal.ETAS.fit(list.input)

  ## Small bit of post processing
  list.output.quietRuninScenario_V2[[i]] <- append(list.input, list(model.fit = ETAS.model.fit))
  list.output.quietRuninScenario_V2[[i]]$runtime <- sum(list.output.quietRuninScenario_V2[[i]]$model.fit$bru_timings$Time)
  list.output.quietRuninScenario_V2[[i]]$nEvents <- length(list.output.quietRuninScenario_V2[[i]]$catalog[,1])
}
```

```{r}
# This version crops out the catalogue subset of interest and is therefore not conditioned on the history

list.output.quietRuninScenario <-list()

  # Load a set of parameters that we will need to tweak for this application

fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
list.input <- create.input.list.temporal.noCatalogue(fpath)
  
  
  ####################
  # Tweak the variables laoded from the input file
list.input$M0 <- M0

i=1
  
  # Change the starting location, measured on the ETAS scale
list.input$mu.init <- startingValues$mu[i]
list.input$alpha.init <- startingValues$alpha[i]
list.input$K.init <- startingValues$K[i]
list.input$c.init <- startingValues$c[i]
list.input$p.init <- startingValues$p[i]

    link.f <- list(mu = \(x) gamma.t(x, a_mu, b_mu), 
                 K = \(x) loggaus.t(x, a_K, b_K), 
                 alpha = \(x) unif.t(x, a_alpha, b_alpha), 
                 c_ = \(x) unif.t(x, a_c, b_c), 
                 p = \(x) unif.t(x, a_p, b_p))
  
  # initial value - convert from ETAS scale to internal scale
  list.input$th.init <- list(th.mu = inv.gamma.t(list.input$mu.init, list.input$a_mu, list.input$b_mu),
                  th.K = inv.loggaus.t(list.input$K.init, list.input$a_K, list.input$b_K),
                  th.alpha = inv.unif.t(list.input$alpha.init, list.input$a_alpha, list.input$b_alpha),
                  th.c = inv.unif.t(list.input$c.init, list.input$a_c, list.input$b_c),
                  th.p = inv.unif.t(list.input$p.init, list.input$a_p, list.input$b_p) )


for (i in 1:length(startTimes)) {  
  startTime <- startTimes[i]
  list.input$time.int <- c(0, modelledDuration-startTime)
  list.input$T12 <- c(0, modelledDuration-startTime)
  
    # Define options for inlabru 
  if(is.null(list.input$max_step)){
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         num.threads = 5,
                         #bru_method = list(max_step = 0.5),
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  } else {
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         bru_method = list(max_step = list.input$max_step),
                         num.threads = 5,
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  }

  ## Add out catalogue to the input list
  list.input$catalog <- catList[[i]]
  
  ## Add the catalogue formatted for bru
  list.input$catalog.bru <- data.frame(
    ts = catList[[i]]$ts,
    magnitudes = catList[[i]]$magnitudes,
    idx.p = seq_len(nrow(catList[[i]]))
  )
  
  ## Input list is now formatted
  ####################  
  
  ## Run the model according to the input list
  ETAS.model.fit <- Temporal.ETAS.fit(list.input)

  ## Small bit of post processing
  list.output.quietRuninScenario[[i]] <- append(list.input, list(model.fit = ETAS.model.fit))
  list.output.quietRuninScenario[[i]]$runtime <- sum(list.output.quietRuninScenario[[i]]$model.fit$bru_timings$Time)
  list.output.quietRuninScenario[[i]]$nEvents <- length(list.output.quietRuninScenario[[i]]$catalog[,1])
}
```

### Plot posteriors with corresponding starting values

```{r fig.height=1.2, fig.width=5}
plots <- list()

trueParas <- data.frame(value=c( mu, K, alpha,c, p), param=c( "mu", "K", "alpha","c", "p"))



post.list <- get_posterior_param(input.list = list.output.quietRuninScenario[[1]])
post.df <-  post.list[[1]]
post.df$id <-startTimes[1]
 
for(i in 2:length(list.output.quietRuninScenario)){
 post.list <- get_posterior_param(input.list = list.output.quietRuninScenario[[i]])
 post.df.tmp <-  post.list[[1]]
 post.df.tmp$id <-startTimes[i]
 
 post.df <- rbind(post.df,post.df.tmp)
 }

plots[[1]] <- ggplot(post.df, aes(x=x,y=y, group=id, color=factor(id), lty=factor(id))) + 
    geom_line() +
  #scale_x_discrete(guide = guide_axis(check.overlap = TRUE)) + 
    facet_wrap(facets = vars(param), scales = 'free', labeller = label_parsed, nrow=1) +
  geom_vline(aes(xintercept = value ), data=trueParas, color='black', linetype=2, label="True value") + 
  labs(color='Start time [days]', lty='Start time [days]' ) +
  ggtitle(paste("(b) Posteriors for a subset catalogue where the intensity is Not conditioned on the history" )) +
  xlab("ETAS Posteriors") +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  theme(legend.position='hidden')

```

```{r fig.height=1.2, fig.width=5}
plots_V2 <- list()

trueParas <- data.frame(value=c( mu, K, alpha,c, p), param=c( "mu", "K", "alpha","c", "p"))

post.list <- get_posterior_param(input.list = list.output.quietRuninScenario_V2[[1]])
post.df <-  post.list[[1]]
post.df$id <-startTimes[1]
 
for(i in 2:length(list.output.quietRuninScenario_V2)){
 post.list <- get_posterior_param(input.list = list.output.quietRuninScenario_V2[[i]])
 post.df.tmp <-  post.list[[1]]
 post.df.tmp$id <-startTimes[i]
 
 post.df <- rbind(post.df,post.df.tmp)
 }

plots_V2[[1]] <- ggplot(post.df, aes(x=x,y=y, group=id, color=factor(id), lty=factor(id))) + 
    geom_line() +
  #scale_x_discrete(guide = guide_axis(check.overlap = TRUE)) + 
    facet_wrap(facets = vars(param), scales = 'free', labeller = label_parsed, nrow=1) +
  geom_vline(aes(xintercept = value ), data=trueParas, color='black', linetype=2, label="True value") + 
  labs(color='Start time [days]', lty='Start time [days]' ) +
  ggtitle(paste("(c) Posteriors of catalogue which is conditioned on the history" )) +
  xlab("ETAS Posteriors") +
    theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1)) #+
  #theme(legend.position='hidden')

```

```{r}
save(list.output.quietRuninScenario, file="list.output.quietRuninScenario.Rda")
save(list.output.quietRuninScenario_V2, file="list.output.quietRuninScenario_V2.Rda")
```


```{r fig.height=3, fig.width=5}
plt <- grid_arrange_shared_legend(catPlot, plots[[1]], plots_V2[[1]], ncol=1, nrow=3, position='bottom')


ggsave("historyResults.png", plt)
ggsave("historyResults.pdf", plt)


```

