---
title: "Modified ETAS"
author: "Farnaz Kamranzad"
date: "2023-02-20"
output: html_document
---


# Calling libraries
```{r, eval=TRUE, warning=FALSE, include=FALSE, echo=FALSE, message=FALSE}
start.time <- Sys.time()
knitr::opts_chunk$set(echo = TRUE)
library(tidyquant)
library(gridExtra)
library(grid)
library(lemon)
library(ggplot2)
library(inlabru)
library(INLA)
library(ETAS.inlabru)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(pracma)
library(Bolstad2)
```



# Calling my .R source file for assessing rate incompleteness
```{r}
# This source features modified functions for considering the rate incompleteness 
source('modified_functions.R')
```



# Making synthetic catalogue 
```{r}
# Defining 'true' ETAS parameters for generating a synthetic catalogue
mu <- 0.1
K <- 0.089
alpha <- 2.29
c <- 0.11
p <- 1.08
M0 <- 2.5
theta <- list(mu=mu, K=K, alpha=alpha, c=c, p=p)

# Defining parameters for modelling incompleteness
G <- 3.8
H <- 1.0
b <- 1.0

# Simulating the synthetics
modelledDuration = 100 
Ht <- data.frame(ts=c(50), magnitudes=c(6.7)) 
samp.etas.list <- generate.temporal.ETAS.synthetic(theta=theta, beta.p=log(10), M0=M0, T1=0, T2=modelledDuration, Ht=Ht)
simulated.ETAS.catalogue <- bind_rows(samp.etas.list)
simulated.ETAS.catalogue <- simulated.ETAS.catalogue[order(simulated.ETAS.catalogue$ts),]
simulated.ETAS.catalogue$idx <- seq(1,length(simulated.ETAS.catalogue$ts))

simulated.ETAS.catalogue
```


# Defining an input list
```{r}
# A dataframe containing different starting values for the inversion algorithm
startingValues <- data.frame( mu=c(0.3),
                              K=c(0.1),
                              alpha=c(1.),
                              c=c(0.2),
                              p=c(1.01))


# Load a set of parameters that we will need to tweak for this application
fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
list.input <- create.input.list.temporal.noCatalogue(fpath)

# Tweak the variables laoded from the input file
list.input$M0 <- M0
list.input$G <- G
list.input$H <- H
list.input$b <- b

i=1

# Change the starting location, measured on the ETAS scale
list.input$mu.init <- startingValues$mu[i]
list.input$alpha.init <- startingValues$alpha[i]
list.input$K.init <- startingValues$K[i]
list.input$c.init <- startingValues$c[i]
list.input$p.init <- startingValues$p[i]

link.f <- list(mu = \(x) gamma_t(x, a_mu, b_mu), 
               K = \(x) loggaus_t(x, a_K, b_K), 
               alpha = \(x) unif_t(x, a_alpha, b_alpha), 
               c_ = \(x) unif_t(x, a_c, b_c), 
               p = \(x) unif_t(x, a_p, b_p))
  
# initial value - convert from ETAS scale to internal scale
list.input$th.init <- list(th.mu = inv_gamma_t(list.input$mu.init, list.input$a_mu, list.input$b_mu),
                           th.K = inv_loggaus_t(list.input$K.init, list.input$a_K, list.input$b_K),
                           th.alpha = inv_unif_t(list.input$alpha.init, list.input$a_alpha, list.input$b_alpha),
                           th.c = inv_unif_t(list.input$c.init, list.input$a_c, list.input$b_c),
                           th.p = inv_unif_t(list.input$p.init, list.input$a_p, list.input$b_p) )

startTime <- 0
list.input$time.int <- c(startTime, modelledDuration)
list.input$T12 <- c(startTime, modelledDuration)


# Define options for inlabru 
if(is.null(list.input$max_step)){
  list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                  bru_max_iter = list.input$max_iter, # maximum number of iterations
                                  num.threads = 5,
                                  #bru_method = list(max_step = 0.5),
                                  inla.mode = 'experimental', # type of inla algorithm
                                  bru_initial = list.input$th.init) # parameters initial values
  } else {
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                                    bru_max_iter = list.input$max_iter, # maximum number of iterations
                                    bru_method = list(max_step = list.input$max_step),
                                    num.threads = 5,
                                    inla.mode = 'experimental', # type of inla algorithm
                                    bru_initial = list.input$th.init) # parameters initial values
    }


#list.input
```



# Attaching the simulated catalogue to the formatted input list
```{r}
## Add out catalogue to the input list
list.input$catalog <- simulated.ETAS.catalogue

## Add the catalogue formatted for bru
list.input$catalog.bru <- data.frame(
  ts = simulated.ETAS.catalogue$ts,
  magnitudes = simulated.ETAS.catalogue$magnitudes,
  idx.p = seq_len(nrow(simulated.ETAS.catalogue))
  )
  
list.input
```



# Primary checks and the incompleteness information
```{r}
# Find the mainshock
Mainshock <- mainshock(list.input)

# Find time interval for rate incompleteness
Other_large_events <- subset(list.input$catalog.bru , magnitudes>5 & magnitudes<Mainshock$magnitudes)

# Attach Mc(t) to the catalog.bru in the input list
list.input <- assign_Mc(list.input, M0, G, H, b)

# time interval between the mainshock and the end of the rate incompleteness
incompleteness_end_time <- Mainshock$ts + 10^((Mainshock$magnitudes-M0-G)/H)

# Check data frequencies
Number_of_all_EQs <- nrow(list.input$catalog.bru)
Number_of_EQs_before_mainshock <- Mainshock$idx.p-1
Number_of_EQs_in_incompletness_interval <- nrow(subset(list.input$catalog.bru , ts>Mainshock$ts & ts<incompleteness_end_time))
Number_of_EQs_to_be_removed <- nrow(subset(list.input$catalog.bru , ts>Mainshock$ts & ts<incompleteness_end_time & magnitudes<Mc))


incompleteness_info <- data.frame(Number_of_all_EQs, Number_of_EQs_before_mainshock, 
                                  Mainshock$ts, incompleteness_end_time,
                                  Number_of_EQs_in_incompletness_interval, Number_of_EQs_to_be_removed)
Other_large_events
incompleteness_info <- as.data.frame(t(incompleteness_info))
incompleteness_info
```



# Calculating the ETAS conditional intensity
```{r, warning=FALSE}
# Defining mh, th, and t
mh <- list.input$catalog.bru$magnitudes
th <- list.input$catalog.bru$ts
t <- seq(0, modelledDuration, length.out=modelledDuration*100)

# Conditional intensity for full catalogue
conditional_intensity <- function (t){cond.lambda(theta=theta, t=t, th=th, mh=mh, M0=M0)}

# Conditional intensity after rate modification
modified_conditional_intensity <- function (t){modified_cond.lambda(theta=theta, t=t, th=th, mh=mh, M0=M0, G=G, H=H, b=b)}

# Values of the functions for a defined sequence of time
cond_vals <- sapply(t , conditional_intensity)
modif_cond_vals <- sapply(t , modified_conditional_intensity)
cond_vals_point <- sapply(60 , conditional_intensity)

#options(max.print=10000)
#cond_vals
```


# Plotting histogram + ETAS conditional intensities
```{r, fig.height=7, fig.width=15}


ggplot() +
  geom_histogram(aes(x=list.input$catalog.bru$ts), col='white', lwd=0.1, fill=rgb(0.1, 0.8, 0.8, 1), bins=200) +
  geom_line(aes(x=t, y=cond_vals, col = "All rates"), lwd=1) +
  geom_line(aes(x=t, y=modif_cond_vals, col = "Modified rates") , lwd=1) +
  #geom_line(aes(x=t, y=testtttt_vals, col = "inc rates") , lwd=1) +
  geom_line(aes(x=t, y=cond_vals_point, col = "Rate at day 60 (example)") , lwd=1) +
  labs(x="Time (day)", y = "Rate", col="Legend") + xlim(0, 100) + ylim(0, 350) + 
  theme_bw() + theme(axis.text=element_text(size=12), axis.title=element_text(size=14, face="bold")) +
  scale_color_manual(values=c("red", "purple", "green", "grey"))

#ggplot() + geom_point(aes(list.input$catalog.bru$ts , list.input$catalog.bru$magnitudes))
```



# Comparing rates with integrals
```{r}
# Analytical integration
T1 <- 0
T2 <- modelledDuration
(T2-T1)*theta$mu + sum(exp((log(K) + alpha*(mh-M0) + log(c/(p-1)) + log(((((max(T1,th)-th)/c)+1)^(1-p))-((((T2-th)/c)+1)^(1-p))))))
(T2-T1)*theta$mu + sum(K * exp(alpha*(mh-M0)) * c/(p-1) *((((max(T1,th)-th)/c)+1)^(1-p)-(((T2-th)/c)+1)^(1-p)))


# Numerical integration
#integrate(modified_conditional_intensity, lower=0, upper=modelledDuration)$value

d_f1 <- data.frame(t , cond_vals, modif_cond_vals)
df_clean <- na.omit(d_f1)

int_all <- trapz(df_clean$t, df_clean$cond_vals)
int_all_sint <- sintegral(df_clean$t, df_clean$cond_vals, n.pts = 256)$int
int_modif <- trapz(df_clean$t, df_clean$modif_cond_vals)
deduction <- int_all-int_modif


# Comparison table
paste('Integral of full function = ' , int_all)
paste('Integral of modified function = ' , int_modif)
paste('Area between the two functions = ' , deduction)
paste('number of all earthquakes = ' , Number_of_all_EQs)
paste('number of removals= ' , Number_of_EQs_to_be_removed)
```



```{r, eval=FALSE}
library(bayesianETAS)
edit(bayesianETAS::sampleETASposterior)
```




# Directly remove data from the incompleteness part (referred as Removed model)
```{r}

list.input_removed <- list.input

simulated.ETAS.catalogue_removed <- data.frame(simulated.ETAS.catalogue)

#t <- seq(0, modelledDuration, 0.1)

for(i in length(simulated.ETAS.catalogue_removed$ts):1 ){
  if(simulated.ETAS.catalogue_removed$ts[i]>Mainshock$ts){ 
    Mc_t <- Mainshock$magnitudes - G - H * log10(simulated.ETAS.catalogue_removed$ts[i]-Mainshock$ts)
  }
  else{Mc_t <- 2.5}

  #Mc_t = max(Mc_t, 2.5, na.rm=TRUE)
  
  if( simulated.ETAS.catalogue_removed$magnitudes[i] < Mc_t){
    simulated.ETAS.catalogue_removed <- simulated.ETAS.catalogue_removed[-c(i),]
  }
}

simulated.ETAS.catalogue_removed$idx <- seq(1,length(simulated.ETAS.catalogue_removed$ts))

list.input_removed$catalog <- simulated.ETAS.catalogue_removed
  

list.input_removed$catalog.bru <- data.frame(
    ts = simulated.ETAS.catalogue_removed$ts,
    magnitudes = simulated.ETAS.catalogue_removed$magnitudes,
    idx.p = seq_len(nrow(simulated.ETAS.catalogue_removed))
  )
```



# Inversion: Fitting ETAS models according to the input lists
```{r}
# Original ETAS fit
ETAS.model.fit_original <- Temporal.ETAS.fit(list.input)

# Removed ETAS fit
ETAS.model.fit_removed <- Temporal.ETAS.fit(list.input_removed)

# Modified ETAS fit
ETAS.model.fit_modified <- modified_Temporal.ETAS.fit(list.input)
```




# Small bit of post processing: Attaching model outputs to the input lists
```{r}
i<-1

# Original Version
list.output_original <-list()
list.output_original[[i]] <- append(list.input, list(model.fit = ETAS.model.fit_original))
list.output_original[[i]]$runtime <- sum(list.output_original[[i]]$model.fit$bru_timings$Time)
list.output_original[[i]]$nEvents <- length(list.output_original[[i]]$catalog[,1])


# Removed Version
list.output_removed <-list()
list.output_removed[[i]] <- append(list.input_removed, list(model.fit = ETAS.model.fit_removed))
list.output_removed[[i]]$runtime <- sum(list.output_removed[[i]]$model.fit$bru_timings$Time)
list.output_removed[[i]]$nEvents <- length(list.output_removed[[i]]$catalog[,1])


# Modified Version
list.output_modified <-list()
list.output_modified[[i]] <- append(list.input, list(model.fit = ETAS.model.fit_modified))
list.output_modified[[i]]$runtime <- sum(list.output_modified[[i]]$model.fit$bru_timings$Time)
list.output_modified[[i]]$nEvents <- length(list.output_modified[[i]]$catalog[,1])
```



# Posterior analysis
```{r, fig.height=4, fig.width=15}
plots <- list()

trueParas <- data.frame(value=c(mu, K, alpha,c, p), param=c("mu", "K", "alpha","c", "p"))

post.list_original <- get_posterior_param(input.list= list.output_original[[1]])
post.df_original <-  post.list_original[[1]]
post.df_original$id <- "Original"

post.list_removed <- get_posterior_param(input.list= list.output_removed[[1]])
post.df_removed <-  post.list_removed[[1]]
post.df_removed$id <- "Removed"

post.list_modified <- get_posterior_param(input.list= list.output_modified[[1]])
post.df_modified <-  post.list_modified [[1]]
post.df_modified$id <- "Modified"

ggplot(NULL, aes(x, y)) + 
  geom_line(data=post.df_removed, aes(group=id, color='Directly Removed'), size=1)+
  geom_line(data=post.df_original, aes(group=id, color='Original'), size=1)+
  geom_line(data=post.df_modified, aes(group=id, color='Modified'), size=1)+
  facet_wrap(facets = vars(param), scales = 'free', labeller = label_parsed, nrow=1) +
  geom_vline(aes(xintercept = value ), data=trueParas, color='black', linetype=2, label="True value") + 
  labs(color='Catalogue scenario', linetype='Catalogue scenario' ) +
  ggtitle(paste("Effect of short term incompleteness after large events on ETAS posteriors" )) +
  xlab("ETAS Posteriors") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))
  #theme(legend.position='hidden')

```



# Computation time
```{r}
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```








