---
title: "temporal_injection_ETAS_dev.Rmd"
author: "Naylor"
date: "2022-10-27"
output: html_document
---

# Results from  Naylor et al.
## Effect of short term incompleteness on ETAS posteriors

In this notebook, we explore the impact of short term catalogue incompleteness after a large event on the accuracy of the posteriors.

- We talk a base line catalogue of 1000 days with a large event on day 500. 
- Then we introduce incompleteness using the formula suggested by Helmstetter et al (). 
- We invert both using inlabru and compare the posteriors with the true values used to generate the synthetic catalogues.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ETAS.inlabru)

library(tidyquant)
library(gridExtra)
library(grid)
library(lemon)
library(ggplot2)

library(inlabru)
library(INLA)

library(dplyr)


```


## Define the parameters for the syntheirc catalogues and starting values for inversion

```{r Create synthetic catalogue and plot}

# Parameters we use to generate synthetics, which we will refer to as the 'true' parameters
mu <- 0.1
K <- 0.089
alpha <- 2.29
c <- 0.11
p <- 1.08

# Format the true ETAS parameters for code to generate the synthetics
theta_etas <- list(mu=mu, K=K, alpha=alpha, c=c, p=p)

# A dataframe containing different starting values for the algorithm
startingValues <- data.frame( mu=c( 0.3),
                              K=c( 0.1),
                              alpha=c( 1.),
                              c=c( 0.2),
                              p=c( 1.01))
nRealisations <- length(startingValues$mu)


# Temporal duration of the synthetic catalogue in days
modelledDuration = 1000  # [days]

# The minimum magnitude that will be modelled
M0 = 2.5

```

## Generate new catalogues and save them

```{r}

############
#### Generate the second catalogue with a M6.7 event on day 500.01

Ht <- data.frame(ts=c(500.0), magnitudes=c(6.7))   # Impose a M6.7 event on day 500.01
samp.etas.list <- generate.temporal.ETAS.synthetic(theta = theta_etas, beta.p = log(10), M0 = M0,  T1 = 0, T2 = modelledDuration, Ht=Ht)

M6p7.ETAS.cat <- bind_rows(samp.etas.list)
M6p7.ETAS.cat <- M6p7.ETAS.cat[order(M6p7.ETAS.cat$ts),]

M6p7.ETAS.cat$idx <- seq(1,length(M6p7.ETAS.cat$ts))

```

```{r}
G <- 3.8
#G <- 4.6
H <- 1.

M6p7.ETAS.cat.incomplete <- data.frame(M6p7.ETAS.cat)

t <- seq(0,1000,0.1)

for(i in length(M6p7.ETAS.cat.incomplete$ts):1 ){
  if(M6p7.ETAS.cat.incomplete$ts[i]>500.){ 
    Mc_t <- 6.7 - G - H * log10(M6p7.ETAS.cat.incomplete$ts[i]-500)
  }
  else{Mc_t <- 2.5}

  #Mc_t = max(Mc_t, 2.5, na.rm=TRUE)
  
  if( M6p7.ETAS.cat.incomplete$magnitudes[i] < Mc_t){
    M6p7.ETAS.cat.incomplete <- M6p7.ETAS.cat.incomplete[-c(i),]
  }
}

M6p7.ETAS.cat.incomplete$idx <- seq(1,length(M6p7.ETAS.cat.incomplete$ts))

```

```{r}

save(M6p7.ETAS.cat,file="M6p7.ETAS.cat.Rda")
save(M6p7.ETAS.cat.incomplete,file="M6p7.ETAS.cat.incomplete.Rda")

```

# Present the catalogues

## Load the catalogues

```{r}
load("catList.Rda")
```

## Plot properties of the catalogues

```{r fig.height=2, fig.width=4}
plots <-list()
 
n <- which(M6p7.ETAS.cat$magnitudes == max(M6p7.ETAS.cat$magnitudes) )

df_large<-data.frame(ts=c(500), magnitudes=c(6.7), n<-c(n))

plots[[1]] <- ggplot() + 
  geom_point(data = M6p7.ETAS.cat, aes(x=ts, y=magnitudes), size=0.1, alpha=0.5) +
  xlim(0, 1000) +
  ggtitle(paste( "A. Complete catalogue, nEvents =", length(M6p7.ETAS.cat$ts) )) +
  ylim(2,7) +
  geom_vline(xintercept=500, lty=2, color='red') +
  geom_point(data=df_large, aes(ts, magnitudes), size=1) +
  theme_bw() +
  xlab("Time [days]") +
  ylab("Magnitude")

   
plots[[2]] <- ggplot() + 
  geom_point(data = M6p7.ETAS.cat.incomplete, aes(x=ts, y=magnitudes), size=0.1, alpha=0.5) +
  xlim(0, 1000) +
  ggtitle(paste( "C. Incomplete catalogue, nEvents =", length(M6p7.ETAS.cat.incomplete$ts) )) +
  ylim(2,7) +
  geom_vline(xintercept=500, lty=2, color='red') +
  geom_point(data=df_large, aes(ts, magnitudes), size=1) +
  theme_bw() +
  xlab("Time [days]") +
  ylab("Magnitude") 


plots[[3]] <- ggplot() + 
  geom_point(data = M6p7.ETAS.cat, aes(x=idx, y=magnitudes), size=0.1, alpha=0.5) +
  ggtitle(paste( "B. Complete" )) +
  ylim(2,7) +
  xlab("Event number") +
  ylab("Magnitude") +
  theme_bw() +
  geom_vline(xintercept = n, lty=2, color='red') +
  geom_point(data=df_large, aes(n, magnitudes), size=1)


plots[[4]] <- ggplot() + 
  geom_point(data = M6p7.ETAS.cat.incomplete, aes(x=idx, y=magnitudes), size=0.1, alpha=0.5) +
  ggtitle(paste( "D. Incomplete" )) +
  ylim(2,7) +
  xlab("Event number") +
  ylab("Magnitude") +
  theme_bw() +
  geom_vline(xintercept = n, lty=2, color='red')+
  geom_point(data=df_large, aes(n, magnitudes), size=1)+
  xlim(0,2000)



#marrangeGrob(plots, nrow=2, ncol=2)

plt <- grid.arrange(plots[[1]],plots[[2]],plots[[3]],plots[[4]], ncol=2, nrow=2, top="Catalogues used for short term incompleteness experiment")

ggsave("cataloguesForIncompletenessExp.png", plt)
ggsave("cataloguesForIncompletenessExp.pdf", plt)
```

# Analyse the complete and incomplete catalogues

## Analysis Complete Catalogue with M6.7 event seeded

```{r}
list.output.incompleteness <-list()

  # Load a set of parameters that we will need to tweak for this application
fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
list.input <- create.input.list.temporal.noCatalogue(fpath)
  
  
  ####################
  # Tweak the variables laoded from the input file
list.input$M0 <- M0

i=1
  
  # Change the starting location, measured on the ETAS scale
list.input$mu.init <- startingValues$mu[i]
list.input$alpha.init <- startingValues$alpha[i]
list.input$K.init <- startingValues$K[i]
list.input$c.init <- startingValues$c[i]
list.input$p.init <- startingValues$p[i]

    link.f <- list(mu = \(x) gamma.t(x, a_mu, b_mu), 
                 K = \(x) loggaus.t(x, a_K, b_K), 
                 alpha = \(x) unif.t(x, a_alpha, b_alpha), 
                 c_ = \(x) unif.t(x, a_c, b_c), 
                 p = \(x) unif.t(x, a_p, b_p))
  
  # initial value - convert from ETAS scale to internal scale
  list.input$th.init <- list(th.mu = inv.gamma.t(list.input$mu.init, list.input$a_mu, list.input$b_mu),
                  th.K = inv.loggaus.t(list.input$K.init, list.input$a_K, list.input$b_K),
                  th.alpha = inv.unif.t(list.input$alpha.init, list.input$a_alpha, list.input$b_alpha),
                  th.c = inv.unif.t(list.input$c.init, list.input$a_c, list.input$b_c),
                  th.p = inv.unif.t(list.input$p.init, list.input$a_p, list.input$b_p) )



  startTime <- 0
  list.input$time.int <- c(startTime, modelledDuration)
  list.input$T12 <- c(startTime, modelledDuration)
  
    # Define options for inlabru 
  if(is.null(list.input$max_step)){
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         num.threads = 5,
                         #bru_method = list(max_step = 0.5),
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  } else {
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         bru_method = list(max_step = list.input$max_step),
                         num.threads = 5,
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  }

  ## Add out catalogue to the input list
  list.input$catalog <- M6p7.ETAS.cat
  
  ## Add the catalogue formatted for bru
  list.input$catalog.bru <- data.frame(
    ts = M6p7.ETAS.cat$ts,
    magnitudes = M6p7.ETAS.cat$magnitudes,
    idx.p = seq_len(nrow(M6p7.ETAS.cat))
  )
  
  ## Input list is now formatted
  ####################  
  
  ## Run the model according to the input list
  ETAS.model.fit <- Temporal.ETAS.fit(list.input)

  i<-1
  ## Small bit of post processing
  list.output.incompleteness[[i]] <- append(list.input, list(model.fit = ETAS.model.fit))
  list.output.incompleteness[[i]]$runtime <- sum(list.output.incompleteness[[i]]$model.fit$bru_timings$Time)
  list.output.incompleteness[[i]]$nEvents <- length(list.output.incompleteness[[i]]$catalog[,1])

```

## Analyse the incomplete version of this catalogue

```{r}

  # Load a set of parameters that we will need to tweak for this application
  fpath <- system.file("extdata", "user_input_synthetic_noCatalogue.txt", package="ETAS.inlabru")
  list.input <- create.input.list.temporal.noCatalogue(fpath)
  
  ####################
  # Tweak the variables laoded from the input file
list.input$M0 <- M0

i=1
  
  # Change the starting location, measured on the ETAS scale
list.input$mu.init <- startingValues$mu[i]
list.input$alpha.init <- startingValues$alpha[i]
list.input$K.init <- startingValues$K[i]
list.input$c.init <- startingValues$c[i]
list.input$p.init <- startingValues$p[i]

    link.f <- list(mu = \(x) gamma.t(x, a_mu, b_mu), 
                 K = \(x) loggaus.t(x, a_K, b_K), 
                 alpha = \(x) unif.t(x, a_alpha, b_alpha), 
                 c_ = \(x) unif.t(x, a_c, b_c), 
                 p = \(x) unif.t(x, a_p, b_p))
  
  # initial value - convert from ETAS scale to internal scale
  list.input$th.init <- list(th.mu = inv.gamma.t(list.input$mu.init, list.input$a_mu, list.input$b_mu),
                  th.K = inv.loggaus.t(list.input$K.init, list.input$a_K, list.input$b_K),
                  th.alpha = inv.unif.t(list.input$alpha.init, list.input$a_alpha, list.input$b_alpha),
                  th.c = inv.unif.t(list.input$c.init, list.input$a_c, list.input$b_c),
                  th.p = inv.unif.t(list.input$p.init, list.input$a_p, list.input$b_p) )



  startTime <- 0
  list.input$time.int <- c(startTime, modelledDuration)
  list.input$T12 <- c(startTime, modelledDuration)
  
    # Define options for inlabru 
  if(is.null(list.input$max_step)){
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         num.threads = 5,
                         #bru_method = list(max_step = 0.5),
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  } else {
    list.input$bru.opt.list <- list(bru_verbose = 3, # type of visual output 
                         bru_max_iter = list.input$max_iter, # maximum number of iterations
                         bru_method = list(max_step = list.input$max_step),
                         num.threads = 5,
                         inla.mode = 'experimental', # type of inla algorithm
                         bru_initial = list.input$th.init) # parameters initial values
  }

  ## Add out catalogue to the input list
  list.input$catalog <- M6p7.ETAS.cat.incomplete
  
  ## Add the catalogue formatted for bru
  list.input$catalog.bru <- data.frame(
    ts = M6p7.ETAS.cat.incomplete$ts,
    magnitudes = M6p7.ETAS.cat.incomplete$magnitudes,
    idx.p = seq_len(nrow(M6p7.ETAS.cat.incomplete))
  )
  
  ## Input list is now formatted
  ####################  
  
  ## Run the model according to the input list
  ETAS.model.fit <- Temporal.ETAS.fit(list.input)

  i<-2
  ## Small bit of post processing
  list.output.incompleteness[[i]] <- append(list.input, list(model.fit = ETAS.model.fit))
  list.output.incompleteness[[i]]$runtime <- sum(list.output.incompleteness[[i]]$model.fit$bru_timings$Time)
  list.output.incompleteness[[i]]$nEvents <- length(M6p7.ETAS.cat.incomplete$catalog[,1])

```
Save the results

```{r}
save(list.output.incompleteness, file="list.output.incompleteness.Rda")
```


# Plot posteriors for the complete and incomplete catalogues

```{r fig.height=1.2, fig.width=5}
plots <- list()

trueParas <- data.frame(value=c( mu, K, alpha,c, p), param=c( "mu", "K", "alpha","c", "p"))


post.list <- get_posterior_param(input.list = list.output.incompleteness[[1]])
post.df <-  post.list[[1]]
post.df$id <- "Complete"
 
for(i in 2:length(list.output.incompleteness)){
 post.list <- get_posterior_param(input.list = list.output.incompleteness[[i]])
 post.df.tmp <-  post.list[[1]]
 post.df.tmp$id <-"Incomplete"
 
 post.df <- rbind(post.df,post.df.tmp)
 }

plots[[1]] <- 
  ggplot(post.df, aes(x=x,y=y, group=id, color=factor(id), lty=factor(id))) + 
    geom_line() +
  #scale_x_discrete(guide = guide_axis(check.overlap = TRUE)) + 
    facet_wrap(facets = vars(param), scales = 'free', labeller = label_parsed, nrow=1) +
  geom_vline(aes(xintercept = value ), data=trueParas, color='black', linetype=2, label="True value") + 
  labs(color='Catalogue scenario', linetype='Catalogue scenario' ) +
  ggtitle(paste("Effect of short term incompleteness after large events on ETAS posteriors" )) +
  xlab("ETAS Posteriors") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust=1))+
  theme(legend.position='hidden')

```


```{r fig.height=1.5, fig.width=5}
plt <- grid_arrange_shared_legend(plots[[1]], ncol=1, nrow=1, position='bottom')

ggsave("posteriorsForIncompletenessExp.png", plt)
ggsave("posteriorsForIncompletenessExp.pdf", plt)
```

# Sample the posteriors of the complete and incomplete inversions and explore the impact on the triggering functions


```{r}
plot_triggering <- list()
plot_triggering[[1]] <- triggering_fun_plot(list.output.incompleteness[[1]], magnitude = 4, n.samp = 100) +
  ggtitle("C. M4 Complete") +     
  theme_bw()  + 
  ylim(0,5.5) + 
  theme(plot.title = element_text(size = 8))

plot_triggering[[2]] <- triggering_fun_plot(list.output.incompleteness[[2]], magnitude = 4, n.samp = 100)+
  ggtitle("D. M4 Incomplete") +     
  theme_bw()  + 
  ylim(0,5.5) + 
  theme(plot.title = element_text(size = 8))

plot_triggering[[3]] <- triggering_fun_plot(list.output.incompleteness[[1]], magnitude = 6.7, n.samp = 100) +
  ggtitle("E. M6.7 Complete") +     
  theme_bw() + 
  ylim(0,1700) + 
  theme(plot.title = element_text(size = 8))

plot_triggering[[4]] <- triggering_fun_plot(list.output.incompleteness[[2]], magnitude = 6.7, n.samp = 100) +
  ggtitle("F. M6.7 Incomplete") +     
  theme_bw() + 
  ylim(0,1700) + 
  theme(plot.title = element_text(size = 8))

```

```{r}
plot_omori <- list()
plot_omori[[1]] <- omori_plot(list.output.incompleteness[[1]], n.samp = 100) +
  ggtitle("A. Omori decay for complete") +     
  theme_bw()  + 
  ylim(0,1) + 
  theme(plot.title = element_text(size = 8))


plot_omori[[2]] <- omori_plot(list.output.incompleteness[[2]], n.samp = 100) +
  ggtitle("B. Omori decay for incomplete")+     
  theme_bw()  + 
  ylim(0,1) + 
  theme(plot.title = element_text(size = 8))

```

```{r}

plt <- grid.arrange(plot_omori[[1]], plot_triggering[[1]], plot_triggering[[3]],plot_omori[[2]],plot_triggering[[2]], plot_triggering[[4]], ncol=3, nrow=2, top="Triggering function variability", left="Incomplete Catalogue              Complete Catalogue")

ggsave("trigFuncVariability_incompleteness.png", plt)
ggsave("trigFuncVariability_incompleteness.pdf", plt)
```
