---
title: "1. Exploratory Data Analysis: Italian Earthquake Catalogue - Horus"
author: "Mark Naylor"
date: "2023-06-27"
output: html_document
---

```{r setup0, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "png",
  dev.args = list(type = "cairo-png"),
  echo = TRUE
)
```

Prior to any statistical analysis, it is important to explore the properties of the dataset. Earthquake catalogues are composed of estimates of the spatial location of events, their data-time, a measure of their magnitude and their depth. The depth is most prone to uncertainty and artefacts, followed by the estimate of the magnitude for which there can be a range of values depending upon the method used to estimate it. Some catalogues also attempt to estimate uncertainty in these parameters.

Below, we provide examples of the sorts of plots you can do to understand important features of catalogue data.
- `ETAS.inlabru` currently requires that a b-value to descrive the frequency-magnitude of earthquakes is provided to the model, so we show how this can be evaluated.
- We have choices in how we subset data for regional events, and we use the example of l'Aquila to show how this can be implemented.

What we don't show in this notebook is how to input the data into a dataframe in the first place. We reccommend using the same column names as below so that it is easy to copy and modify the code.

# Setup

## Load Packages

```{r setup, include=TRUE,message=FALSE,warning=FALSE}
library(ETAS.inlabru)
library(ggplot2)

library(rnaturalearth)
library(terra)
library(sf)
library(ggspatial)
library(rnaturalearthdata)
library(dplyr)
library(lubridate)

## This is just the EPSG equivalent of WGS84
crs_wgs84 <- st_crs("EPSG:4326")
```

# Load an earthquake catalogue

Earthquake data is stored in the so-called earthquake catalogues. Many different catalogues exists for the same region and there is no easy way to decide which one is better. Here, we provide the HOmogenized instRUmental Seismic (HORUS) catalogue from 1960 to 2020. It can be downloaded from [http://horus.bo.ingv.it/](http://horus.bo.ingv.it/).
A subset (see [data-raw/horus.R](https://github.com/edinburgh-seismicity-hub/ETAS.inlabru/blob/main/R/AnalysingPosteriors.R) for how the the subset was created)
of the information in the Horus catalogue can be loaded using

```{r}
# load HORUS catalogue
data(horus, package = "ETAS.inlabru")

# transform time string in Date object
horus$time_date <- as.POSIXct(
  horus$time_string,
  format = "%Y-%m-%dT%H:%M:%OS"
)

head(horus)

# Set df_cat to be the catalogue we want to analyse and then use this below
df_cat <- horus

# Add an integer id to each event
df_cat$event_num <- seq.int(nrow(df_cat))

# Generate an sf version of the catalogue where the longitude and latitude are
# converted to a point object
df_cat.sf <- st_as_sf(df_cat,
  coords = c("lon", "lat"),
  crs = crs_wgs84
)

head(df_cat.sf)
```

The data reports for each earthquake the time as a string (`time_string`),
the longitude (`lon`) and latitude (`lat`) of the epicentre, the depth in
kilometres (`depth`), and the moment magnitude (`M`).   

## Get map and extract crs for the map

We download a polygon file for the national boundary of Italy for use later in this notebook.

```{r}
italy.map <-
  ne_countries(country = "Italy", returnclass = "sf", scale = "medium")
italy.crs <- crs(italy.map)
print(italy.crs)

ggplot() +
  geom_sf(data = italy.map) +
  theme_bw() +
  ggtitle("Map of the Italian territorial boundary and coastline")
```


# Exploratory Data Analysis on the whole of Italy using the Horus catalogue

## Histogram of event depths

The histogram of depth data with 1km bins initially looks reasonable except for the artefact of a cluster of events at zero depth.

```{r}
hist(
  df_cat$depth,
  breaks = seq(650, -5, -1),
  xlim = c(-5, 40),
  main = "Histogram of depths, 1km bins",
  xlab = "Depth [km]"
)
```

However, when we reduce the bin width to 100m we see that there are more artefacts in the data. The spikes at 0, 5, 10 and 11km are artefacts of the depth estimation, one cause is that that the algorithms may have got stuck at their starting values.

```{r}
hist(
  df_cat$depth,
  breaks = seq(650, -5, -0.1),
  xlim = c(-5, 40),
  main = "Histogram of depths (100m bins)",
  xlab = "Depth [km]"
)
```
The magnitude time series shows how the completeness of intermediate and deep events has improved over time. 
```{r}
ggplot(df_cat, aes(x = time_date, y = depth)) +
  geom_point(size = 0.1) +
  ylim(620, 0) +
  ggtitle("Depth timeseries using natural time")
```
  
## Plot of event magnitudes using natural time

Natural time refers to where we plot each event evenly spaced along the x-axis. The advantage of this is that data that occurs in periods of high activity are spread out, which means that we can see structure in the data when the rates are high.

Below we plot the magnitude of the events on the y-axis and natural time on the horizontal.

Some of the observations are:

- There are over 40,000 events with magnitude data
- There is a cluster of events with $M=0$ which is likely to be an artefact of how the catalogue is generated
- There are period of time when the data are all rounded to 1 decimal place producing horizontal banding
- At other times, some of the data is not rounded
- The sensitivity at the lowest magnitudes varies in time as can be seen by the changes in the lowest magnitudes

```{r}
ggplot(df_cat, aes(x = event_num, y = M)) +
  geom_point(size = 0.1) +
  ggtitle("Magnitude timeseries using natural time")
```

## Plot of event magnitudes using date-time

Here we present the same data as in the previous plot, but plotted against the data-time. Clustered events now sit on top of each other. We get a better representaion of how the data density increases towards the modern day and how the magnitude incompleteness varyies in time.

```{r Magnitude natural-time series}
ggplot(df_cat, aes(x = time_date, y = M)) +
  geom_point(size = 0.1) +
  ggtitle("Horus catalogue magnitude time series")
```

The number of events occuring each year is difficult to assess in this plot, so below we group the data by year and count the number of events occuring each year. The plot of this shows how the number of events being recorded each year is gradually increasing. This artefact is due to the gradual increase in the sensitivity of the network, not that the seismicity is systematically increasing.

```{r}
df_cat$Time <- as.Date(df_cat$time_date)

print("Extract year")
# extract the year and convert to numeric format
df_cat$Year <- as.numeric(format(df_cat$Time, "%Y"))

countEventsInYear <- df_cat %>%
  group_by(Year) %>%
  summarize(counts = n())

plot(
  countEventsInYear,
  type = "l",
  main = "Annual number of events for whole catalogue"
)
```


```{r Magnitude time series}
# Filtered for M>4
ggplot(df_cat[df_cat$M > 4, ], aes(x = time_date, y = M)) +
  geom_point(size = 0.1) +
  ggtitle("Whole catalogue magnitude timeseries for M>4")
```

```{r}
countEventsInYear <- df_cat[df_cat$M >= 4, ] %>%
  group_by(Year) %>%
  summarize(counts = n())

plot(countEventsInYear, type = "l", main = "Annual number of events with M>=4")
abline(h = mean(countEventsInYear), col = 2)
```

## Frequency-magnitude analysis for Horus catalogue

The traditional GR plot shows cumulative data (red) and we also show the incremental data (black). I personally prefer looking at the incremental data because it is not correlated and therefore easier to understand how Poisson counting errors account for uncertainty in the data.

There are three domains in a standard GR plot.
- At low magnitudes there will always be incompleteness due to limitations of the monitoring network
- A requirement of GR being a good model is that there is a log-linear portion of the data where the slope is stable, this slope is refered to as the b-value
- At high magnitudes there will always be fluctuations in the magnitudes due to the effect of counting errors on bins with low rates

```{r National GR Plot}
minMag <- 1.5
maxMag <- max(df_cat$M)

mags <- df_cat[df_cat$M > minMag, ]$M

tmp <- hist(mags, breaks = seq(minMag - 0.05, maxMag + 0.1, 0.1), plot = FALSE)

N.counts <- length(tmp$counts)
tmp$cumulativeCounts <- cumsum(tmp$counts[N.counts:1])[N.counts:1]

m.min <- 4
bin_m.min <- which(tmp$mids == m.min)
freq_m.min <- tmp$counts[bin_m.min]
b <- 1
x <- tmp$mids
y <- freq_m.min * 10^(-b * (x - m.min))
y.cum <- tmp$cumulativeCounts[bin_m.min] * 10^(-b * (x - m.min))

ggplot() +
  geom_point(aes(x = tmp$mids, y = tmp$counts)) +
  geom_point(aes(x = tmp$mids, y = tmp$cumulativeCounts),
    color = "red",
    pch = "+",
    size = 2
  ) +
  scale_y_log10() +
  ggtitle(paste("Frequency-magnitude plot with arbitary GR dist: b =", b)) +
  xlab("Magnitude") +
  ylab("log10(Frequency)") +
  geom_line(aes(x = x, y = y)) +
  geom_line(aes(x = x, y = y.cum), color = "red") +
  geom_vline(xintercept = m.min, lty = 2)
```

## b-value stability plot for the events in the Horus catalogue

```{r National b-value stability}
minMag <- 2
maxMag <- max(df_cat$M, na.rm = TRUE)
mags <- df_cat[df_cat$M >= minMag, ]$M

x <- seq(minMag, maxMag, 0.1)

b.stability.list <- c()
b.error.list <- c()
m.mean <- c()

b_utsu <- c()
b_guttorp <- c()
b_elst <- c()
delta_b_utsu <- c()
b_elst_lower <- c()
b_elst_upper <- c()

max.index.x <- length(x) - 5

for (i in 1:max.index.x) {
  mag.threshold <- x[i]
  mags.subset <- mags[mags > mag.threshold]

  N <- length(mags.subset)
  b_utsu[i] <- 1 / (log(10) * (mean(mags.subset) - mag.threshold + 0.05))
  delta_b_utsu[i] <- log(10) * b_utsu[i]**2 *
    sqrt(sum((mags.subset - mean(mags.subset))**2) / (N * (N - 1)))

  b_guttorp[i] <- 1 / (2 * 0.05 * log(10)) *
    log((mean(mags.subset) - mag.threshold + 2 * 0.05) /
      (mean(mags.subset) - mag.threshold))

  deltaMags <- diff(mags.subset)
  deltaMags_p <- deltaMags[deltaMags > 0.1]
  N <- length(deltaMags_p)
  b_elst[i] <- 1 / (2 * 0.05 * log(10)) *
    log((mean(deltaMags_p)) / (mean(deltaMags_p) - 0.1))
  c <- 10**(0.1 * b_elst[i])
  b_elst_lower[i] <- 1 / (0.1 * log(10)) *
    log((c + sqrt(c / N)) / (1 + sqrt(c / N)))
  b_elst_upper[i] <- 1 / (0.1 * log(10)) *
    log((c - sqrt(c / N)) / (1 - sqrt(c / N)))
}

ggplot() +
  geom_line(aes(x = x[1:max.index.x], y = b_utsu)) +
  geom_line(
    aes(x = x[1:max.index.x], y = b_utsu + delta_b_utsu),
    color = 2,
    lty = 2
  ) +
  geom_line(
    aes(x = x[1:max.index.x], y = b_utsu - delta_b_utsu),
    color = 2, lty = 2
  ) +
  geom_line(
    aes(x = x[1:max.index.x], y = b_guttorp),
    color = 3, lty = 1
  ) +
  xlab("Magnitude threshold") +
  ylab("b-value estimate") +
  geom_hline(yintercept = 1, lty = 3) +
  geom_hline(yintercept = 0.85, lty = 3) +
  ggtitle("b-value stability plot for Horus catalogue")
```

## Heatmap of the events in the Horus catalogue

```{r hexbin event density}
ggplot() +
  geom_hex(data = df_cat[df_cat$M > 3, ], aes(x = lon, y = lat), bins = 50) +
  scale_fill_continuous(type = "viridis") +
  geom_sf(
    data = italy.map,
    fill = alpha("lightgrey", 0),
    color = "orange",
    size = 0.2
  ) +
  ggtitle("Density plot for M>3 events") +
  theme_bw()
```

## Map of the events in the Italian Horus catalogue

Here we plot the location of the seismicity on a map. The outline of the Italy territory is shown in green.


```{r map of locations}
ggplot() +
  geom_sf(data = df_cat.sf[df_cat$M > 3, ], size = 0.05) +
  geom_sf(
    data = italy.map,
    fill = alpha("lightgrey", 0),
    color = "green",
    linewidth = 0.7
  ) +
  geom_sf(data = df_cat.sf[df_cat$M > 5, ], size = 0.5, color = "orange") +
  geom_sf(data = df_cat.sf[df_cat$M > 6, ], size = 0.5, color = "red") +
  ggtitle("Map of event locations")
```

```{r}
eventDate <- ymd_hms("2009-04-06 00:00:00")
endDate <- eventDate + days(400)
startDate <- eventDate - days(50)
deltaLat <- 2.4
latLims <- c(42.2, 42.5)
longLims <- c(13, 13.75)

ggplot() +
  geom_point(data = df_cat[df_cat$M > 3, ], aes(time_date, lat), size = 0.1) +
  geom_point(
    data = df_cat[df_cat$M > 5, ],
    aes(time_date, lat),
    size = 1.2,
    color = "orange"
  ) +
  geom_point(
    data = df_cat[df_cat$M > 6, ],
    aes(time_date, lat),
    size = 1.5,
    color = "red"
  ) +
  ggtitle("Italian latitude-time plot") +
  geom_rect(
    aes(
      xmin = as.POSIXct(startDate),
      xmax = as.POSIXct(endDate),
      ymin = latLims[1],
      ymax = latLims[2]
    ),
    alpha = 0.4,
    fill = "blue",
    color = "blue"
  )
```

# Analysis if the L'Aquila Sequence

To focus on the L Aquila seismic sequence is sufficient to retain only the observations in a specific space-time-magnitude region that include the sequence of interest. For the L'Aquila sequence, we retain all the events with magnitude greater or equal than $2.5$ happened during 2009 with longitude in $(10.5, 16)$ and latitude in $(40.5, 45)$. The L'Aquila sequence selected in this way should be composed by 1024 events. Any other seismic sequence can be selected similarly.

To do the selection is convenient to transform the time string in a `Date` object and to select the rows of the Horus catalogue verifying the conditions.

start.date = "2009-04-06 00:00:00 BST"
end.date = "2010-01-01 00:00:00 BST"
magnitude.completeness = 2.49
min.longitude = 13.00
max.longitude = 13.75
min.latitude = 42.2
max.latitude = 42.5 

```{r Extract l Aquila subset}
eventDate <- ymd_hms("2009-04-06 00:00:00")
endDate <- eventDate + days(400)
startDate <- eventDate - days(50)
deltaLat <- 2.4
latLims <- c(42.2, 42.5)
longLims <- c(13, 13.75)

minMAG <- 0

# Subset the main catalogue
df_cat.subset <- df_cat[df_cat$M >= minMAG, ]
df_cat.subset <- df_cat.subset[(df_cat.subset$lat > latLims[1]), ]
df_cat.subset <- df_cat.subset[(df_cat.subset$lat < latLims[2]), ]
df_cat.subset <- df_cat.subset[(df_cat.subset$lon > longLims[1]), ]
df_cat.subset <- df_cat.subset[(df_cat.subset$lon < longLims[2]), ]

head(df_cat.subset)
```
## Plot of the event magnitude time series: l'Aquila

```{r}
ggplot() +
  geom_point(
    data = df_cat.subset[df_cat.subset$M > 3, ], aes(time_date, lat), size = 0.1
  ) +
  geom_point(
    data = df_cat.subset[df_cat.subset$M > 5, ],
    aes(time_date, lat),
    size = 1.2,
    color = "orange"
  ) +
  geom_point(
    data = df_cat.subset[df_cat.subset$M > 6, ],
    aes(time_date, lat),
    size = 1.5,
    color = "red"
  ) +
  ggtitle("L'Aquila latitude-time plot") +
  geom_rect(
    aes(
      xmin = as.POSIXct(startDate),
      xmax = as.POSIXct(endDate),
      ymin = latLims[1],
      ymax = latLims[2]
    ),
    alpha = 0.4, fill = "blue", color = "blue"
  )
```

```{r regional GR Plot}
minMag <- 0
maxMag <- max(df_cat.subset$M, na.rm = TRUE)

mags <- df_cat.subset[df_cat.subset$M >= minMag, ]$M

tmp <- hist(mags, breaks = seq(minMag - 0.05, maxMag + 0.4, 0.1), plot = FALSE)

N.counts <- length(tmp$counts)
tmp$cumulativeCounts <- cumsum(tmp$counts[N.counts:1])[N.counts:1]

m.min <- 4
bin_m.min <- which(tmp$mids == m.min)
freq_m.min <- tmp$counts[bin_m.min]
b <- 1.
x <- tmp$mids
y <- freq_m.min * 10^(-b * (x - m.min))
y.cum <- tmp$cumulativeCounts[bin_m.min] * 10^(-b * (x - m.min))

ggplot() +
  geom_point(aes(x = tmp$mids, y = tmp$counts)) +
  geom_point(
    aes(x = tmp$mids, y = tmp$cumulativeCounts),
    color = "red",
    pch = "+"
  ) +
  scale_y_log10() +
  ggtitle(paste("Frequency-magnitude plot with arbitary GR dist: b =", b)) +
  xlab("Magnitude") +
  ylab("log10(Frequency)") +
  geom_line(aes(x = x, y = y)) +
  geom_line(aes(x = x, y = y.cum), color = "red") +
  geom_vline(xintercept = m.min, lty = 2)
```

## b-value stability plot: l'Aquila

```{r}
minMag <- 0
maxMag <- max(df_cat.subset$M, na.rm = TRUE)
mags <- df_cat.subset[df_cat.subset$M >= minMag, ]$M

x <- seq(minMag, maxMag, 0.1)

b.stability.list <- c()
b.error.list <- c()
m.mean <- c()

b_utsu <- c()
b_guttorp <- c()
b_elst <- c()
delta_b_utsu <- c()
b_elst_lower <- c()
b_elst_upper <- c()

max.index.x <- length(x) - 5

for (i in 1:max.index.x) {
  mag.threshold <- x[i]
  mags.subset <- mags[mags > mag.threshold]

  N <- length(mags.subset)
  b_utsu[i] <- 1 / (log(10) * (mean(mags.subset) - mag.threshold + 0.05))
  delta_b_utsu[i] <- log(10) * b_utsu[i]**2 *
    sqrt(sum((mags.subset - mean(mags.subset))**2) / (N * (N - 1)))

  b_guttorp[i] <- 1 / (2 * 0.05 * log(10)) *
    log((mean(mags.subset) - mag.threshold + 2 * 0.05) /
      (mean(mags.subset) - mag.threshold))

  deltaMags <- diff(mags.subset)
  deltaMags_p <- deltaMags[deltaMags > 0.1]
  N <- length(deltaMags_p)
  b_elst[i] <- 1 / (2 * 0.05 * log(10)) *
    log((mean(deltaMags_p)) / (mean(deltaMags_p) - 0.1))
  c <- 10**(0.1 * b_elst[i])
  b_elst_lower[i] <- 1 / (0.1 * log(10)) *
    log((c + sqrt(c / N)) / (1 + sqrt(c / N)))
  b_elst_upper[i] <- 1 / (0.1 * log(10)) *
    log((c - sqrt(c / N)) / (1 - sqrt(c / N)))
}

ggplot() +
  geom_line(aes(x = x[1:max.index.x], y = b_utsu)) +
  geom_line(
    aes(x = x[1:max.index.x], y = b_utsu + delta_b_utsu),
    color = 2, lty = 2
  ) +
  geom_line(
    aes(x = x[1:max.index.x], y = b_utsu - delta_b_utsu),
    color = 2, lty = 2
  ) +
  geom_line(
    aes(x = x[1:max.index.x], y = b_guttorp),
    color = 3, lty = 1
  ) +
  xlab("Magnitude threshold") +
  ylab("b-value estimate") +
  geom_hline(yintercept = 1, lty = 3) +
  geom_hline(yintercept = 0.85, lty = 3) +
  ggtitle("b-value stability plot for the l'Aquila catalogue")
```

